---
title: "MSU Campus Trees - Proposed Nature's Notebook Sites"
format: html
editor: visual
page-layout: full
---

Below is a simple interactive map with locations of trees observed on the MSU campus between 2017 and 2025. Proposed sites for Nature's Notebook are outlined in blue (with red ID numbers near the center). Trees that were observed in 2025 are shown as filled circles, and trees that were not observed in 2025 are shown with open circles. Hovering over a point will display the tree species and accession number. Clicking on a point will display that information, plus the years that tree was monitored and the total number of observations submitted. A summary of trees in each proposed site is below the map.

```{r gather-summarize-data}
#| echo: FALSE
#| output: FALSE
#| warning: FALSE

library(dplyr)
library(leaflet)
library(htmltools)
library(terra)
library(tidyterra)
library(sf)
library(kableExtra)

# Load tree list (and remove any trees without lat/lon)
trees <- read.csv("data/MSUTreeList.csv")
trees <- trees %>%
  mutate(common_name = factor(common_name)) %>%
  filter(!is.na(latitude))

# Load phenology data, 2017-2025
dat <- read.csv("data/msu-phenology-data-2017-2025.csv")

# Determine which years each tree was monitored and attach to tree list
dat <- dat %>%
  group_by(tree) %>%
  summarize(minyr = min(year),
            maxyr = max(year),
            nyrs = n_distinct(year),
            y2025 = ifelse(2025 %in% year, 1, 0),
            nobs = n()) %>%
  mutate(yrs = paste0(minyr, "-", maxyr)) %>%
  mutate(obs = paste0(nyrs, " (", nobs, ")")) %>%
  data.frame() %>%
  select(tree, yrs, obs, y2025)
trees <- trees %>%
  left_join(dat, by = c("accession" = "tree"))

# Load proposed sites and names
sites <- vect("nn-sites/nn-sites-revised.shp")
sitenames <- read.csv("nn-sites/nn-site-names-revised.csv")

# Get centroid locations for labeling
sites_centroid <- terra::centroids(sites) %>%
  as.data.frame(geom = "XY") %>%
  left_join(select(sitenames, -current_name), by = c("site_no" = "number")) %>%
  rename(sitename = new_name)

# Attach site to trees dataframe
treesv <- vect(trees, geom = c("longitude", "latitude"), crs = "epsg:4326")
trees_sites <- terra::extract(sites, treesv)
trees$site <- trees_sites$name
trees$site_no <- trees_sites$site_no
trees <- left_join(trees, select(sitenames, -current_name), 
                   by = c("site_no" = "number")) %>%
  rename(sitename = new_name)

# Summarize trees at each site
sites_summary <- trees %>%
  group_by(site_no, sitename) %>%
  summarize(n_trees = n(),
            n_spp = n_distinct(common_name),
            n_trees_2025 = sum(y2025),
            n_spp_2025 = n_distinct(common_name[y2025 == 1]),
            .groups = "drop") %>%
  data.frame()
```

```{r map}
#| echo: FALSE

# Convert SpatVector to sf
sites_sf <- sf::st_as_sf(sites)

# Map settings
pal_leaf <- leaflet::colorFactor(palette = "turbo", 
                                 domain = unique(trees$common_name))

treespp <- sort(unique(trees$common_name))
cols <- pal_leaf(treespp)

# Create map
trees %>%
  leaflet(height = 750) %>%
  addProviderTiles(providers$OpenStreetMap,
                   options = providerTileOptions(opacity = 0.6)) %>% 
  addCircleMarkers(
    data = filter(trees, y2025 == 0),
    lng = ~longitude,
    lat = ~latitude,
    label = ~paste0(common_name, ": ", accession),
    color = ~pal_leaf(common_name),
    fillColor = ~pal_leaf(common_name),
    fillOpacity = 0,
    stroke = TRUE,
    weight = 2,
    radius = 5,
    popup = ~paste0(common_name, "<br>",
                    "ID: ", accession, "<br>", 
                    "Site number: ", site_no, "<br>",
                    "Years: ", yrs, "<br>",
                    "No. years (no. obs): ", obs)
  ) %>%
  addCircleMarkers(
    data = filter(trees, y2025 == 1),
    lng = ~longitude,
    lat = ~latitude,
    label = ~paste0(common_name, ": ", accession),
    fillColor = ~pal_leaf(common_name),
    fillOpacity = 0.7,
    color = "black",
    stroke = TRUE,
    weight = 2,
    radius = 5,
    popup = ~paste0(common_name, "<br>",
                    "ID: ", accession, "<br>",
                    "Site number: ", site_no, "<br>",
                    "Years: ", yrs, "<br>",
                    "No. years (no. obs): ", obs)
  ) %>%
  addLegend(
    position = "bottomleft",
    pal = pal_leaf,
    values = ~ common_name,
    labels = ~common_name,
    title = "Tree Species"
  ) %>%
  addPolylines(data = sites_sf, 
               weight = 3) %>%
  addLabelOnlyMarkers(data = sites_centroid, 
                      lat = ~y,
                      lng = ~x,
                      label = ~sitename,
                      labelOptions = labelOptions(
                        noHide = TRUE,
                        direction = "center",
                        textOnly = TRUE,
                        textsize = "14px",
                        style = list("font-weight" = "bold",
                                     "color" = "black"))) %>%
  addScaleBar(
    position = "topright"
  )
```

\

```{r table}
#| echo: FALSE

sites_summary %>%
  kable(col.names = c("Site number", "Name", "No. trees", "No. species", "No. trees", "No. species"),
        align = c("c", "r", rep("c", 4))) %>%
  add_header_above(header = c(" " = 2, "Total" = 2, "Observed in 2025" = 2))

```
